Approximately Exact Calculations for Linear Mixed Models

### 2015-05-29 ms.tex
RLL analysis of HMO

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
\begin{equation*}
	\maxit=10; \hspace{1cm} \epsilon=5; \hspace{1cm}
	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} M=5
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 10: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 10 iterations there is a total of 640 boxes, which are displayed in Figure~\ref{fig:hmo1}.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 8 & 2 & -1618.84\\
3 & 16 & 6 & -1509.24\\
4 & 32 & 14 & -1408.12\\
5 & 44 & 35 & -1325.99\\
6 & 56 & 65 & -1265.01\\
7 & 104 & 95 & -1256.35\\
8 & 184 & 153 & -1243.51\\
9 & 224 & 281 & -1240.60\\
10 & 180 & 460 & -1239.49\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
\label{table:hmo_HH11_run}
\end{table}

@@

Figure~\ref{fig:hmoboxes} shows the outlines of the 640 boxes.  The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.  Figure~\ref{fig:hmorll} shows the same boxes, shaded by $L^b$ in each box.  The red dot is the lower left corner of the box that maximizes $L^b$.  Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.  
Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.  So we rerun the algorithm on that box and with more stringent control parameters:
\begin{equation*}
	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
\end{equation*}
Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.  For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-1320.51$\\
2 & 16 & 0 & -1276.23\\
3 & 64 & 0 & -1252.18\\
4 & 256 & 0 & -1243.52\\
5 & 1024 & 0 & -1239.61\\
6 & 4020 & 19 & -1237.78\\
7 & 13128 & 757 & -1236.70\\
8 & 14092 & 10362 & -1236.16\\
9 & 9444 & 22093 & -1235.88\\
10 & 5716 & 30108 & -1235.88\\
11 & 2256 & 35260 & -1235.88\\
12 & 0 & 37516 & -1235.88\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
\label{table:hmo_HH11_run2}
\end{table}

\begin{figure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
	\caption{Locations of the boxes.}
	\label{fig:hmoboxes2}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
	\label{fig:hmorll2}
  \end{subfigure}
  \caption{The 37,516 boxes produced in the second run of the algorithm for the
	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
  \label{fig:hmo2}
\end{figure}

Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.  Figure~\ref{fig:hmorll2} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll2} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll2} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.5$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.  Repeat as needed.

>>>>

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
\begin{equation*}
	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} \text{ratio $=$ TRUE}; \hspace{1cm} M=5
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 15 iterations there is a total of 895 boxes, which are displayed in Figure~\ref{fig:hmo1}.

%\begin{table}[h]
%\centering
%\begin{tabular}{|c|c|c|c|}
%\hline
%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
%1 & 4 & 0 & $-\infty$\\
%2 & 8 & 2 & -1618.84\\
%3 & 16 & 6 & -1509.24\\
%4 & 32 & 14 & -1408.12\\
%5 & 44 & 35 & -1325.99\\
%6 & 56 & 65 & -1265.01\\
%7 & 104 & 95 & -1256.35\\
%8 & 184 & 153 & -1243.51\\
%9 & 224 & 281 & -1240.60\\
%10 & 180 & 460 & -1239.49\\
%\hline
%\end{tabular}
%\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
%\label{table:hmo_HH11_run}
%\end{table}

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 8 & 2 & -1618.84\\
3 & 12 & 7 & -1509.24\\
4 & 12 & 16 & -1408.12\\
5 & 12 & 25 & -1325.99\\
6 & 16 & 33 & -1265.01\\
7 & 20 & 44 & -1256.35\\
8 & 28 & 57 & -1243.51\\
9 & 48 & 73 & -1240.60\\
10 & 76 & 102 & -1239.49\\
11 & 132 & 145 & -1239.40\\
12 & 232 & 219 & -1239.40\\
13 & 420 & 346 & -1239.40\\
14 & 172 & 723 & -1239.40\\
15 & 0 & 895 & -1239.40\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 15 iterations for the HMO data.}
\label{table:hmo_HH11_run}
\end{table}

@@

Figure~\ref{fig:hmoboxes} shows the outlines of the 895 boxes.  The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.  Figure~\ref{fig:hmorll} shows the same boxes, shaded by $L^b$ in each box.  The red dot is the lower left corner of the box that maximizes $L^b$.  Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.  
%Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.  So we rerun the algorithm on that box and with more stringent control parameters:
%\begin{equation*}
%	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
%\end{equation*}
%Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.

%\begin{table}[h]
%\centering
%\begin{tabular}{|c|c|c|c|}
%\hline
%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
%1 & 4 & 0 & $-1320.51$\\
%2 & 16 & 0 & -1276.23\\
%3 & 64 & 0 & -1252.18\\
%4 & 256 & 0 & -1243.52\\
%5 & 1024 & 0 & -1239.61\\
%6 & 4020 & 19 & -1237.78\\
%7 & 13128 & 757 & -1236.70\\
%8 & 14092 & 10362 & -1236.16\\
%9 & 9444 & 22093 & -1235.88\\
%10 & 5716 & 30108 & -1235.88\\
%11 & 2256 & 35260 & -1235.88\\
%12 & 0 & 37516 & -1235.88\\
%\hline
%\end{tabular}
%\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
%\label{table:hmo_HH11_run2}
%\end{table}

%\begin{figure}
%  \begin{subfigure}{.5\textwidth}
%	\centering
%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
%	\caption{Locations of the boxes.}
%	\label{fig:hmoboxes2}
%  \end{subfigure}
%  \begin{subfigure}{.5\textwidth}
%	\centering
%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
%	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
%	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
%	\label{fig:hmorll2}
%  \end{subfigure}
%  \caption{The 37,516 boxes produced in the second run of the algorithm for the
%	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
%	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
%  \label{fig:hmo2}
%\end{figure}

%Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.
Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.5$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

%The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.  Repeat as needed.
